---
title: "Easy Stretch of Fixtures"
output: html_document
---

It's a pain to identify stretches of appealing fixtures. The coloured indicators that FPL tacked on in 2016-17 don't seem to update with performances.

The charts below try to highlight four aspects that one would typically look for when trying to decide transfers.

The thresholds, etc. aren't some sort of recommendation. Feel free to change them if some other number makes more sense.

PS1: I might update this every week or not. 
PS2: I think this code should be okay with double gameweeks. If it isn't, 'll try and fix it at the earliest.

```{r Libraries, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

vcLibraries = c('ggplot2','data.table','RCurl','rjson','XML','zoo')

vcInstallLibraries = setdiff(
   vcLibraries, 
   installed.packages()
)
   
if ( length(vcInstallLibraries) > 0 ) {
   install.packages(vcInstallLibraries)
}

lapply(
   vcLibraries, 
   require, 
   character.only = TRUE
)

theme_set(theme_bw(12))

```

```{r GlobalParameters, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# The minimum length of the rolling window to be marked as a stretch.
iGWSpan = 6
# The minimum number of matches that must meet the crtieria in any rolling window to be marked as an easy stretch
iNbrEasyMatches = 4

dtTeamID = data.table(
   TeamID = 1:20,
   Team = c(
      'ARS',
      'BOU',
      'BUR',
      'CHE',
      'CRY',
      'EVE',
      'HUL',
      'LEI',
      'LIV',
      'MCI',
      'MUN',
      'MID',
      'SOU',
      'STK',
      'SUN',
      'SWA',
      'TOT',
      'WAT',
      'WBA',
      'WHU'
   )
)

```

```{r FunctionDeclarations, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

#' Identify whether a criterion is met a certain number of times in a continuous stretch of gameweeks
#' @param viGW integer vector of gameweeks
#' @param vOpponentStat vector of relevant statistic
#' @param iStatThreshold The threshold to check against vOpponentStat
#' @param cThresholdOperator Limited to operators. Applied as cThresholdOperator(vOpponentStat, iStatThreshold)
#' @param iGWStretch Number of gameweeks to include in rolling window
#' @param iNbrEasyMatches The minimum number (i.e. >=) of matches that must satisfy the iStatThreshold for that window to be marked TRUE.
#' @return boolean vector against viGW to mark the weeks which meet the criteria
fEasyMatchesInStretch = function(
   viGW,
   vOpponentStat,
   iStatThreshold,
   cThresholdOperator = '>',
   iGWStretch = iGWSpan,
   iNbrEasyMatches = iNbrEasyMatches
) {

   # sorting the statistic in order of gameweeks
   vOpponentStat = vOpponentStat[order(viGW)]
   viUnsortedGW = viGW
   viGW = sort(viGW)

   # Checking the criterion in a rolling window of length iGWStretch
   dtEasyStretches = data.table(
      EasyStretch = sapply(
         (min(viGW) - iGWStretch + 1):max(viGW),
         function(iGW) {

            sum(
               get(cThresholdOperator)(
                  vOpponentStat[viGW %in% c(iGW:(iGW + iGWStretch - 1))],
                  iStatThreshold
               )
            ) >= iNbrEasyMatches


         }
      ),
      GW = (min(viGW) - iGWStretch + 1):max(viGW)
   )

   # Same GW might be part of multiple windows. 
   # Aggretating this information to check whether a GW is a part of any window at all
   dtEasyStretches = dtEasyStretches[,
      list(EasyStretch = any(EasyStretch)),
      GW
   ]

   # If no easy streches, then return all FALSE.
   if (dtEasyStretches[!is.na(EasyStretch), !any(EasyStretch)]) {
      return ( rep(F, length(viGW)) )
   }

   # Else return TRUEs against the GWs which are a part of an easy stretch
   dtEasyStretches = dtEasyStretches[
      EasyStretch == T,
      list(GW2 = c((GW:(GW+iGWStretch-1)))),
      GW
   ]

   viUnsortedGW %in% dtEasyStretches[, GW2]

}

```

```{r LoadingData, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# scraping data
dtData = data.table()
lData = list()

for ( iGW in 1:38 ) {

   jsonGW <- readLines(
      paste0('https://fantasy.premierleague.com/drf/fixtures/?event=', iGW)
   )
   
   lGW <- fromJSON(
      paste(
         jsonGW, 
         collapse=""
      )
   )
   
   lData[[length(lData) + 1]] = lGW

   dtData = rbind(
      dtData,
      rbindlist(
         lapply(
            lGW,
            function( lGWMatch ) {

               rbind(
                  data.table(
                     TeamID = lGWMatch$team_h,
                     OpponentID = lGWMatch$team_a,
                     Scored = ifelse(
                        is.null(lGWMatch$team_h_score),
                        NA,
                        lGWMatch$team_h_score
                     ),
                     Conceded = ifelse(
                        is.null(lGWMatch$team_a_score),
                        NA,
                        lGWMatch$team_a_score
                     ),
                     At = 'H'
                  ),
                  data.table(
                     TeamID = lGWMatch$team_a,
                     OpponentID = lGWMatch$team_h,
                     Scored = ifelse(
                        is.null(lGWMatch$team_a_score),
                        NA,
                        lGWMatch$team_a_score
                     ),
                     Conceded = ifelse(
                        is.null(lGWMatch$team_h_score),
                        NA,
                        lGWMatch$team_h_score
                     ),
                     At = 'A'
                  )
               )
            }
         )
      )[, GW := iGW],
      fill = T
   )

}

dtData = merge(
   dtData,
   dtTeamID,
   'TeamID'
)

dtData = merge(
   dtData,
   dtTeamID[, list(OpponentID = TeamID, Opponent = Team)],
   'OpponentID'
)

# Finding the last GW that each team has played till
dtData[,
   PlayedGW := max(GW[!is.na(Scored)]),
   Team
]

# Assigning points for the results
dtData[
   Scored > Conceded, 
   Points := 3L
]

dtData[
   Scored == Conceded, 
   Points := 1L
]

dtData[
   Scored < Conceded, 
   Points := 0L
]

# Cumulatively calculating points, average conceded, average scored, etc.
# Doing this not from the start of the season but from iGWSpan number of
# gameweeks from the last played gameweek
setkey(dtData, Team, GW)
dtData[
   GW > PlayedGW - iGWSpan & GW <= PlayedGW, 
   AveragePoints := cumsum(Points) / c(1:.N), 
   list(Team)
]

dtData[
   GW > PlayedGW - iGWSpan & GW <= PlayedGW, 
   AverageScored := cumsum(Scored) / c(1:.N), 
   list(Team)
]

dtData[
   GW > PlayedGW - iGWSpan & GW <= PlayedGW, 
   AverageConceded := cumsum(Conceded) / c(1:.N), 
   list(Team)
]

dtData[
   !is.na(AveragePoints), 
   Rank := 21L - rank(AveragePoints, ties = 'max'), 
   GW
]

# The number at the last played GW is the number we consider for the future
dtData[,
   AveragePoints := na.locf(AveragePoints, na.rm = F), 
   Team
]

dtData[, 
   AverageScored := na.locf(AverageScored, na.rm = F), 
   Team
]

dtData[,
   AverageConceded := na.locf(AverageConceded, na.rm = F), 
   Team
]

dtData[,
   Rank := na.locf(Rank, na.rm = F), 
   Team
]

# Getting the above details for the opponents on that GW
dtData = merge(
   dtData,
   dtData[
      GW == PlayedGW, 
      list(
         Opponent = Team,
         OpponentRank = Rank,
         OpponentAveragePoints = AveragePoints,
         OpponentAverageScored = AverageScored,
         OpponentAverageConceded = AverageConceded
      )
   ],
   c('Opponent')
)

# The merge adds this data to the historic GWs also. Which is wrong.
# Removing it
dtData[ 
   GW <= PlayedGW, 
   OpponentRank := NA
]

dtData[ 
   GW <= PlayedGW, 
   OpponentAveragePoints := NA
]

dtData[ 
   GW <= PlayedGW, 
   OpponentAverageScored := NA
]

dtData[ 
   GW <= PlayedGW, 
   OpponentAverageConceded := NA
]

setkey(dtData, Team, GW)

```


```{r PointForm, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 10, fig.height = 8}

iStatThreshold = 10
iNbrEasyMatches = 5
iGWStretch = iGWSpan
cThresholdOperator = '>='

# Identifying easy stretches based on the table standings from the last iGWSpan gameweeks
dtEasyStretchesFuture = dtData[
   GW > PlayedGW,
   list(
      GW = GW[
         fEasyMatchesInStretch(
            viGW = GW,
            vOpponentStat = OpponentRank,
            iStatThreshold = iStatThreshold,
            cThresholdOperator = cThresholdOperator,
            iGWStretch = iGWStretch,
            iNbrEasyMatches = iNbrEasyMatches
         )
      ]
   ),
   Team
]

setkey(dtEasyStretchesFuture, Team, GW)
dtEasyStretchesFuture[,
   Stretch := cumsum(c(2, diff(GW))>1),
   Team
]

dtEasyStretchesFuture = merge(
   dtData[, 
      list(GW, Team, Rank)
   ],
   dtEasyStretchesFuture,
   c('GW','Team')
)

# Doing the plot

ggplot(dtData[GW > PlayedGW]) +
   geom_line(
      data = dtEasyStretchesFuture, 
      aes(
         x = GW, 
         y = Team, 
         group = paste(Stretch, Team), 
         alpha = -Rank
      ),
      size = 5
   ) +
   geom_point(
      aes(
         x = GW, 
         y = Team, 
         fill = OpponentRank
      ), 
      shape = 21, 
      size = 5
   ) +
   scale_fill_continuous(
      low = 'red', 
      high = 'green'
   ) +
   scale_alpha_continuous(
      breaks = c(-20,-10,-1), 
      labels = c(20, 10, 1), 
      name = 'Team Rank'
   ) +
   theme(
      legend.position = 'bottom',
      legend.background = element_rect(colour = 'black'),
      plot.background = element_rect(fill = "grey95")
   ) +
   labs(
      title = 'Favourable Stretches of Fixtures Against Low Ranking Teams',
      caption = paste0(
         "\nAt least ", iNbrEasyMatches, " matches in ", iGWSpan," gameweeks with opponents ranked ", cThresholdOperator, " ", iStatThreshold, "
         The black streaks mark the easy stretches. The darker it is, the higher the rank of the team itself in recent weeks.
         Author: TheComeOnMan
         Data: premierleague.com"
      ),
      x = 'Game week',
      y = 'Team'
   )

```


```{r ScoredForm, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 10, fig.height = 8}

iStatThreshold = 1
iNbrEasyMatches = 4
cThresholdOperator = '<='
iGWStretch = iGWSpan
# Identifying easy stretches based on the goals scored by the oppostion from the last iGWSpan gameweeks
dtEasyStretchesFuture = dtData[
   GW > PlayedGW,
   list(
      GW = GW[
         fEasyMatchesInStretch(
            viGW = GW,
            vOpponentStat = OpponentAverageScored,
            cThresholdOperator = cThresholdOperator,
            iStatThreshold = 1,
            iGWStretch = iGWStretch,
            iNbrEasyMatches = iNbrEasyMatches
         )
      ]
   ),
   Team
]

setkey(dtEasyStretchesFuture, Team, GW)
dtEasyStretchesFuture[, Stretch := cumsum(c(2, diff(GW))>1), Team]
dtEasyStretchesFuture = merge(
   dtData[GW >= PlayedGW, list(GW, Team, AverageConceded)],
   dtEasyStretchesFuture,
   c('GW','Team')
)

# Doing the plot

ggplot(dtData[GW > PlayedGW]) +
   geom_line(data = dtEasyStretchesFuture, aes(x = GW, y = Team, group = paste(Stretch, Team), alpha = -AverageConceded), size = 5) +
   geom_point(aes(x = GW, y = Team, fill = OpponentAverageScored), shape = 21, size = 5) +
   ggtitle('Favourable stretches to buy defensive players') +
   # scale_color_manual(values = c('H' = 'black', 'A' = 'grey80'), name = 'Played At') +
   scale_fill_continuous(low = 'green', high = 'red', name = 'Average Goals Scored\nBy Opponent Recently') +
scale_alpha_continuous(breaks = c(-2,-1.5,-1,0), labels = c(2, 1.5, 1, 0), name = 'Average Goals\nConceded Recently') +
   theme(
      legend.position = 'bottom',
      legend.background = element_rect(colour = 'black'),
      plot.background = element_rect(fill = "grey95")
   ) +
   labs(
      title = 'Favourable Stretches of Fixtures to Buy Defensive Players',
      caption = paste0("\nAt least ", iNbrEasyMatches," matches in ", iGWSpan," gameweeks with opponents who have scored ", cThresholdOperator, " ", iStatThreshold, " goals on average in the last ", iGWSpan," matches.
       The black streaks mark the easy stretches. The darker it is, the better the defending results for the team itself in recent weeks.
      Author: TheComeOnMan
      Data: premierleague.com"),
      x = 'Game week',
      y = 'Team'
   )

```


```{r ConcededForm, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 10, fig.height = 8}

iStatThreshold = 1.75
iNbrEasyMatches = 4
cThresholdOperator = '>='
iGWStretch = iGWSpan
# Identifying easy stretches based on the goals conceded by the oppostion from the last iGWSpan gameweeks
dtEasyStretchesFuture = dtData[
   GW > PlayedGW,
   list(
      GW = GW[
         fEasyMatchesInStretch(
            viGW = GW,
            vOpponentStat = OpponentAverageConceded,
            cThresholdOperator = cThresholdOperator,
            iStatThreshold = iStatThreshold,
            iGWStretch = iGWStretch,
            iNbrEasyMatches = iNbrEasyMatches
         )
      ]
   ),
   Team
]

setkey(dtEasyStretchesFuture, Team, GW)
dtEasyStretchesFuture[, Stretch := cumsum(c(2, diff(GW))>1), Team]
dtEasyStretchesFuture = merge(
   dtData[GW >= PlayedGW, list(GW, Team, AverageScored)],
   dtEasyStretchesFuture,
   c('GW','Team')
)

# Doing the plot
ggplot(dtData[GW > PlayedGW]) +
   geom_line(data = dtEasyStretchesFuture, aes(x = GW, y = Team, group = paste(Stretch, Team), alpha = AverageScored), size = 5) +
   geom_point(aes(x = GW, y = Team, fill = OpponentAverageConceded), shape = 21, size = 5) +
   # scale_color_manual(values = c('H' = 'black', 'A' = 'grey80'), name = 'Played At') +
   scale_fill_continuous(low = 'red', high = 'green', name = 'Average Goals Conceded\nBy Opponent Recently') +
scale_alpha_continuous(breaks = c(2,1.5,1,0), labels = c(2, 1.5, 1, 0), name = 'Average Goals\nScored Recently') +
   theme(
      legend.position = 'bottom',
      legend.background = element_rect(colour = 'black'),
      plot.background = element_rect(fill = "grey95")
   ) + 
   labs(
      title = 'Favourable Stretches of Fixtures to Buy Offensive Players',
      caption = paste0("\nAt least ", iNbrEasyMatches," matches in ", iGWStretch," gameweeks with opponents who have conceded ", cThresholdOperator, " ", iStatThreshold, " goals on average in the last ", iGWSpan," matches.
      The black streaks mark the easy stretches. The darker it is, the better the attacking results for the team itself in recent weeks.
      Author: TheComeOnMan
      Data: premierleague.com"),
      x = 'Game week',
      y = 'Team'
   )


if ( F ) {
 
   ggplot(dtData[GW > PlayedGW]) +
      geom_line(data = merge(dtData[GW >= PlayedGW, list(GW, Team, AverageScored, OpponentAverageConceded)],dtEasyStretchesFuture,c('GW','Team')), aes(x = GW, y = OpponentAverageConceded, group = paste(Stretch, Team), alpha = AverageScored), size = 3) +
      geom_line(aes(x = GW, y = OpponentAverageConceded), size = 0.5) +
      geom_point(aes(x = GW, y = OpponentAverageConceded, color = At), size = 3) +
      scale_color_manual(values = c('H' = 'black', 'A' = 'grey80'), name = 'Played At') +
      scale_fill_continuous(low = 'red', high = 'green', name = 'Average Goals Conceded\nBy Opponent Recently') +
   scale_alpha_continuous(breaks = c(2,1.5,1,0), labels = c(2, 1.5, 1, 0), name = 'Average Goals\nScored Recently') +
      ggtitle('Favourable Stretches of Fixtures to Buy Offensive Players') +
      theme(
         legend.position = 'bottom',
         legend.background = element_rect(colour = 'black')
      ) +
      facet_grid(Team~.)
   
}


```

```{r HomeGames, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 10, fig.height = 8}

iStatThreshold = 'H'
iNbrEasyMatches = 3
cThresholdOperator = '=='
iGWStretch = 4

# Identifying easy stretches based on the table standings from the last iGWSpan gameweeks
dtEasyStretchesFuture = dtData[
   GW > PlayedGW,
   list(
      GW = GW[
         fEasyMatchesInStretch(
            viGW = GW,
            vOpponentStat = At,
            iStatThreshold = iStatThreshold,
            cThresholdOperator = cThresholdOperator,
            iGWStretch = iGWStretch,
            iNbrEasyMatches = iNbrEasyMatches
         )
      ]
   ),
   Team
]

setkey(dtEasyStretchesFuture, Team, GW)
dtEasyStretchesFuture[,
   Stretch := cumsum(c(2, diff(GW))>1),
   Team
]

dtEasyStretchesFuture = merge(
   dtData[GW >= PlayedGW, list(GW, Team, Rank)],
   dtEasyStretchesFuture,
   c('GW','Team')
)

# Doing the plot

ggplot(dtData[GW > PlayedGW]) +
   geom_line(data = dtEasyStretchesFuture, aes(x = GW, y = Team, group = paste(Stretch, Team), alpha = -Rank), size = 5) +
   geom_point(aes(x = GW, y = Team, fill = OpponentRank), shape = 21, size = 5) +
   scale_fill_continuous(low = 'red', high = 'green') +
   scale_alpha_continuous(breaks = dtEasyStretchesFuture[, -rev(sort(unique(Rank)))], labels = dtEasyStretchesFuture[, rev(sort(unique(Rank)))], name = 'Team Rank') +
   theme(
      legend.position = 'bottom',
      legend.background = element_rect(colour = 'black'),
      plot.background = element_rect(fill = "grey95")
   ) +
   labs(
      title = 'Favourable Stretches of Fixtures Being Played At Home',
      caption = paste0("\nAt least ", iNbrEasyMatches," matches over ", iGWStretch," gameweeks with the match at ", iStatThreshold, "
      The black streaks mark the stretches with lots of home matches. The darker it is, the higher the rank of the team itself in recent weeks.
      Author: TheComeOnMan
      Data: premierleague.com"),
      x = 'Game week',
      y = 'Team'
   )

```